Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models

### 2017-12-14 Paper_v7.tex
Working on LATeX issues

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{subfig}
%\usepackage[space]{cite}
\urlstyle{rm}

@@

%\def\includegraphic{}
%\def\includegraphics{}

@@

\title{Parameter estimation in biochemical models}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={asg242@cornell.edu}   % email address
]{\fnm{Adithya} \snm{Sagar}}
\author[
   addressref={aff1},
   email={ril34@cornell.edu},
   noteref={n1}
]{\fnm{Rachel} \snm{LeCover}}
\author[
   addressref={aff1},
   corref={aff1},
   email={jdv27@cornell.edu}
]{\inits{D}\fnm{Jeffrey} \snm{Varner}}
\author[
   addressref={aff2},
   email={jdv27@cornell.edu}
]{\fnm{Christine} \snm{Shoemaker}}

@@

\subsection*{Phase switching was critical to DOPS performance.}
A differentiating feature of DOPS is the switch to dynamically dimensioned search following stagnation of the initial particle swarm phase.
We quantified the influence of the number of sub-swarms and the switch to DDS on error convergence by comparing DOPS with and without DDS for different numbers of sub-swarms (Fig. \ref{fig-sub-swarms}).
We considered multi swarm particle swarm optimization with and without the DDS phase for $\mathcal{N}$ = 4000 function evaluations and $\mathcal{T}$ = 25 trials on the coagulation model.
We used one, two, four, five and eight sub-swarms, with a total of 40 particles divided evenly amongst the swarms.
Hence, we did not consider swarm numbers of three and seven.
All other algorithm parameters remained the same for all cases.
Generally, the higher sub-swarm numbers converged in fewer function evaluations, where the optimum particle partitioning was in the neighborhood of five sub-swarms.
However, the difference in convergence rate was qualitatively similar for four, five and eight sub-swarms, suggesting there was an optimal number of particles per swarm beyond
which there was no significant advantage.
The multi-swarm particle swarm optimization stagnated after 25\% of the available function evaluations irrespective of the number of sub-swarms.
However, DOPS (with five sub-swarms) switched to DDS after detecting the stagnation.
The DDS phase refined the globally best particle to produce significantly lower error on average when compared to multi-swarm particle swarm optimization alone.
Thus, the automated switching strategy was critical to the overall performance of DOPS.

\subsection*{Continued Switching May Be Beneficial}
We explored the performance of DOPS if it was permitted to switch between PSO and DDS modes. This mode ('Experimental DOPS') performed comparably to DOPS on 10-d Ackley and Rastrigin as well as on 300-dimensional Rastrigin. It performed better than DOPS on the CHO metabolism problem (Fig. \ref{fig-exp-DOPS-b4}a), with the average functional value for Experimental DOPS being nearly half of that of DOPS. To further distinguish DOPS from Experimental DOPS, we compared the performance of each algorithm on the Eggholder function, which is considered to be a difficult function to optimize, owing to its multiple minima. On this function, Experimental DOPS outperformed DOPS however, neither version of DOPS reached the true minimum at -959.6407 on any trial with a budget of 4,000 function evaluations (Fig. \ref{fig-exp-DOPS-b4}b). We also explored the performance of both versions of DOPS on the 100 dimensional Styblinksi-Tang function \cite{jamil2013literature} (Fig. \ref{fig-exp-DOPS-b4}c). In this comparison, experimental DOPS far outperforms DOPS, succeeding in finding the true minimum before exhausting its budget of functional evaluations, whereas DOPS does not reach the minimum. Since the performance of Experimental DOPS was promising on these problems, we decided to measure its performance on the coagulation problem. At 4,000 function evaluations, DOPS performed better than Experimental DOPS, however, the difference between the two method's final average functional values was small (Fig. \ref{fig-exp-DOPS-b4}d), with the final average objective value for DOPS reaching .9413\% of the initial functional value, compared to .9428\% for Experimental DOPS.

@@

\subsection*{Experimental DOPS}
We sought to investigate if switching methods more than once would result in better performance. Experimental DOPS begins with the PSO phase and uses the same criteria as DOPS to switch to the DDS phase. The technique switches back to PSO phase when the DDS phase has reduced the functional value to 90\% of its initial value. Should the DDS phase fail to improve the functional value sufficiently, this version is identical to DOPS, but we found that for the CHO problem as well as for coagulation, the DDS phase did produce sufficient improvement in functional value, resulting in a switch back to the PSO phase from the DDS phase. When the switch from DDS to PSO occurs, we use the best solution from DDS to seed the particle swarm.

DOPS and Experimental DOPS source code is available for download under a MIT license at http://www.varnerlab.org.

@@

@@

\section*{Author's contributions}
    A.S. originally developed algorithm. R.L developed experimental DOPS and generated all of the figures. JDV assisted in manuscript preparation.

\section*{Acknowledgements}
 This study was supported by an award from the Army Research Office (ARO \#59155-LS).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{./figs/Figure_1_Algorithm.pdf}
\caption{\csentence{Schematic of the dynamic optimization with particle swarms (DOPS) approach.} \textbf{A}: Each particle represents an N dimensional parameter vector. Particles are given randomly generated initial solutions and grouped into different sub-swarms. Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience. After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms. When the error due to the global best particle (best particle amongst all the sub-swarms) does not drop over a certain number of function evaluations, the swarm search is stopped and the search switches to a Dynamically Dimensioned Search with global best particle as the initial solution vector or candidate vector.
\textbf{B}: The candidate vector performs a greedy global search for the remaining number of function evaluations. The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step. The probability that a dimension is perturbed decreases as the number of function evaluations increase.
}\label{fig-algorithm}
\end{figure}

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}

% Imports required for p-code of JuPOETs
\usepackage[linesnumbered]{algorithm2e}
\usepackage{courier}
%\usepackage{algorithm}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

@@

\def\includegraphic{}
\def\includegraphics{}

@@

\title{Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={asg242@cornell.edu}   % email address
]{\fnm{Adithya} \snm{Sagar}}
\author[
   addressref={aff1},
   email={ril34@cornell.edu},
   noteref={n1}
]{\fnm{Rachel} \snm{LeCover}}
\author[
   addressref={aff2},
   email={jdv27@cornell.edu}
]{\fnm{Christine} \snm{Shoemaker}}
\author[
   addressref={aff1},
   corref={aff1},
   email={jdv27@cornell.edu}
]{\inits{D}\fnm{Jeffrey} \snm{Varner}}

@@

\subsection*{Phase switching was critical to DOPS performance.}
A differentiating feature of DOPS is the switch to dynamically dimensioned search following stagnation of the initial particle swarm phase.
We quantified the influence of the number of sub-swarms and the switch to DDS on error convergence by comparing DOPS with and without DDS for different numbers of sub-swarms (Fig. \ref{fig-sub-swarms}).
We considered multi swarm particle swarm optimization with and without the DDS phase for $\mathcal{N}$ = 4000 function evaluations and $\mathcal{T}$ = 25 trials on the coagulation model.
We used one, two, four, five and eight sub-swarms, with a total of 40 particles divided evenly amongst the swarms.
Hence, we did not consider swarm numbers of three and seven.
All other algorithm parameters remained the same for all cases.
Generally, the higher sub-swarm numbers converged in fewer function evaluations, where the optimum particle partitioning was in the neighborhood of five sub-swarms.
However, the difference in convergence rate was qualitatively similar for four, five and eight sub-swarms, suggesting there was an optimal number of particles per swarm beyond
which there was no significant advantage.
The multi-swarm particle swarm optimization stagnated after 25\% of the available function evaluations irrespective of the number of sub-swarms.
However, DOPS (with five sub-swarms) switched to DDS after detecting the stagnation.
The DDS phase refined the globally best particle to produce significantly lower error on average when compared to multi-swarm particle swarm optimization alone.
Thus, the automated switching strategy was critical to the overall performance of DOPS. However, it was unclear if multiple strategy switches could further improve perfomance.

We explored the performance of DOPS if it was permitted to switch between PSO and DDS modes multiple times.
This mode (\texttt{Experimental~DOPS}) performed comparably to DOPS on 10-d Ackley and Rastrigin as well as on 300-dimensional Rastrigin function.
\texttt{Experimental~DOPS} performed better than DOPS on the CHO metabolism problem (Fig. \ref{fig-exp-DOPS-b4}a),
with the average functional value being nearly half of that of DOPS.
To further distinguish DOPS from \texttt{Experimental~DOPS}, we compared the performance of each algorithm on the Eggholder function,
a difficult function to optimize given its its multiple minima.
On the Eggholder function, \texttt{Experimental~DOPS} outperformed DOPS however, neither version of DOPS reached the true minimum at -959.6407 on any trial with a budget of 4,000 function evaluations (Fig. \ref{fig-exp-DOPS-b4}b).
We also explored the performance of both versions of DOPS on the 100 dimensional Styblinksi-Tang function \cite{jamil2013literature} (Fig. \ref{fig-exp-DOPS-b4}c).
In this comparison, \texttt{Experimental~DOPS} significantly outperformed DOPS, finding the true minimum before exhausting its functional evaluation budget, while DOPS does not reach the minimum.
Since the performance of \texttt{Experimental~DOPS} was promising on these problems, we measured its performance on the coagulation problem.
Surprisingly, DOPS performed slightly better than \texttt{Experimental~DOPS} on the coagulation problem (Fig. \ref{fig-exp-DOPS-b4}d);
the final average objective value for DOPS reached 0.9413\% of the initial functional value, compared to 0.9428\% for \texttt{Experimental~DOPS}.

@@

\subsection*{Experimental DOPS}
We investigated where switching methods more than once would result in better performance; this DOPS variant is referred to as \texttt{Experimental~DOPS}.
\texttt{Experimental~DOPS} begins with the PSO phase and uses the same criteria as DOPS to switch to the DDS phase. The technique switches back to PSO phase when the DDS phase has reduced the functional value to 90\% of its initial value.
Should the DDS phase fail to improve the functional value sufficiently, this version is identical to DOPS, but we found that for the CHO problem as well as for coagulation, the DDS phase did produce sufficient improvement in functional value, resulting in a switch back to the PSO phase from the DDS phase. When the switch from DDS to PSO occurs, we use the best solution from DDS to seed the particle swarm.
DOPS and Experimental DOPS source code is available for download under a MIT license at http://www.varnerlab.org.

@@

\section*{Ethics approval and consent to participate}
Not applicable

\section*{Consent to publish}
Not applicable

@@

\section*{Author's contributions}
    A.S. developed original DOPS algorithm. R.L developed experimental DOPS and generated all of the figures.
    CS developed the dynamic dimensional search approach, and assisted in editing the manuscript.
    JDV directed the study, prepared and edited the manuscript.

\section*{Availability of data and materials}
DOPS is open source, available under an MIT software license.
The DOPS source code is freely available from the DOPS GitHub repository at https://github.com/varnerlab/POETs.jl.
All samples used in this study are included in the \texttt{sample/biochemical} and \texttt{sample/test\_functions} subdirectories of the JuPOETs GitHub repository.

\section*{Funding}
This study was supported by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\clearpage

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.85\textwidth]{./figs/Figure_1_Algorithm.pdf}
\caption{\csentence{Schematic of the dynamic optimization with particle swarms (DOPS) approach.} \textbf{A}: Each particle represents an N dimensional parameter vector.
Particles are given randomly generated initial solutions and grouped into different sub-swarms. Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience. After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms. When the error due to the global best particle (best particle amongst all the sub-swarms) does not drop over a certain number of function evaluations, the swarm search is stopped and the search switches to a Dynamically Dimensioned Search with global best particle as the initial solution vector or candidate vector.
\textbf{B}: The candidate vector performs a greedy global search for the remaining number of function evaluations. The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
The probability that a dimension is perturbed decreases as the number of function evaluations increase.
}\label{fig-algorithm}
\end{figure}
